### 2020-07-05 chapters/background.tex
Multiple grammar corrections

Batch processing is the processing of data grouped by batches. A batch is a collection of data points that have been grouped by certain criteria. Examples of such criteria could be \textit{"all the transactions from the past two months"}, \textit{"the last million transactions"} or even \textit{"all the transactions made by an user until now"}. Batch datasets are known as static data \cite{Martin-Batch-Defin}. An example application of such a technique would be processing all the keywords searched for in the google search engine, and then grouping the results by keyword to count how many times a keyword was used in searches (for trend analysis purposes). Between batch and stream processing, batch processing is the older paradigm. Historically, Apache Hadoop \cite{borthakur2007hadoop, Hadoop, ApacheHadoop} was one of the first widely known frameworks for large-scale batch processing, created in 2005. Hadoop uses a programming model called MapReduce \cite{MapReduce}. This paradigm applies a map function that processes key/value pairs and generates sets of intermediate ones, ultimately merging all intermediate values
using a reducer function. To do this, the distributed computation framework reads and writes from disk on intermediate steps, which makes it inefficient for applications that reuse working sets of data across multiple parallel operations (\textit{e.g.} iterative Machine Learning algorithms). Nearly a decade later, Apache Spark \cite{ApacheSpark, Spark} was released as the industry accepted successor.
Spark retains the scalability and fault tolerance of MapReduce but introduces Resilient Distributed Datasets (RDDs) \cite{SparkRDDs}. An RDD is a read-only collection of objects partitioned across a set of machines that can be rebuilt if a partition is lost.
Additionally, Spark processes data in random access memory (RAM), while Hadoop MapReduce persists data to the disk after a map or reduce action. Thus, Spark achieves lower processing times.

@@

% boundless aggregations no need for windows
Computations over an unbounded dataset are possible. For example, computing the \textit{average} of an event's field for all incoming events is computationally trivial: we need only to store the total number of events and the current total sum of all events arriving in the system. However, is this boundless average over the stream of events useful? While in some scenarios it may be, more often than not we would like to introduce some boundaries into our computations. For instance, consider the monitoring of machines spread out across multiple racks in a large data center and its temperatures. Assume that temperatures measured across all racks are sent to a monitoring system in the form of a data stream. The average temperature of a machine over an infinite period (\textit{i.e.} since the machine was deployed) is of little value for detecting if the equipment is currently overheating, and to trigger the appropriate action to prevent a failure. In this scenario, an average of the temperatures per rack and for the last 5 minutes allows for a more actionable context, in this case identifying a likely-to-fail rack. To obtain such insight requires computing the average over the last 5 minutes' worth of data, effectively defining a begin and end --- \textit{i.e.} all the data between the current timestamp in minutes \textit{t} and \textit{t-5}. By creating such a conceptual range, we realize that in practice we need a bounded dataset. Partitioning an infinite data stream to obtain such a finite dataset is made through the use of windows.

Botan et. al \cite{Botan-SECRET} define a window as:
\begin{definition}
A window W over a stream $\mathbb{S}$ is a finite subset of $\mathbb{S}$
\end{definition}
Windows can be categorized by two dimensions: the window size $W$ and the sliding step size $S$. Windows can be defined as tuple or time-based. Tuple-based windows are windows whose size is defined by \emph{number} of items contained in the window, whereas time-based windows are windows whose size is defined by the total \emph{time} a window can store. For instance, a tuple-based window of size $10^3$ tuples will store exactly $10^3$ items. In contrast, a time-based window storing the last 5 minutes of events can contain a varying number of tuples but will always keep the events arrived in the last 5 minutes. A window slides over incoming data and groups it, enabling us to compute aggregations over a defined and bounded set of events. The step size \textit{S} of a windowing technique is defined as the distance between consecutive windows \cite{Botan-SECRET}. This means that two consecutive windows W\textsubscript{1} and W\textsubscript{2} are exactly \textit{S} units away. As mentioned, these units can be units of time or a number of tuples.

Windows can also be characterized by how they move across data, \textit{i.e.}, how the step $S$ is defined. The three most common ways are: true sliding, stepping or tumbling windows. A true sliding window is a window with a unitary step, i.e., where $S = 1$. At each step, the window advances one unit, which may be one event for tuple-based steps, or one time-unit whether that is in milliseconds, seconds, days, weeks or another magnitude of time, measured based on the system clock or event time. Such a true sliding movement can be observed in Figure \ref{fig:sliding-window} with a tuple-based window of size \textit{W} of three tuples and unitary step.

@@

For the scope of this Thesis, we will work under a tuple-based true sliding window framework. However, despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and other step values as well.

@@

While some aggregations can be computed on an infinite dataset, often we want to compute these over a finite one. As we have seen in Section \ref{sec:windows}, we can obtain a bounded dataset from a data stream by applying windowing techniques. When doing so, we obtain a finite dataset denoted as a sliding window and compute aggregations over it. Tangwongsan \emph{et. al} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs) \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}. As an example of a SWAG, consider the data center temperature monitoring problem from Section \ref{sec:windows}. In this example, we wanted to compute the average temperature per rack in the data center in the last 5 minutes. If we translate this SWAG into streaming SQL following a similar syntax to the one provided by Apache Flink \cite{ApacheFlink} we would get:

@@

In the query above, the
\texttt{TUMBLE(timestamp, Time.minutes(5), rack\_id)} statement groups incoming data using time-based tumbling windows of 5 minutes duration, using the \texttt{timestamp} field to evict events not in this time interval. The resulting list of rows has \texttt{rack\_id} as its key and as corresponding value a list of temperatures, from the past 5 minutes. Finally, for each rack and its list of temperatures, we compute an \texttt{AVG} aggregation, resulting in the average temperature per rack in the past 5 minutes. Note that this is very similar to any aggregation made in a standard Database Management System (DBMS). The main difference is that in standard SQL the \texttt{GROUP BY} clause includes only the columns on which the data will be grouped while in streaming SQL the \texttt{GROUP BY} also specifies the type of window used (\textit{e.g.} in this case, a tumbling window).

@@

\begin{algorithm}
    \caption{Recalculate-From-Scratch insert, evict and query methods}
    \label{pseudo:rfs}
    \begin{algorithmic}[1]
        \Function{query}{vals}
            \State $agg\gets\overline{\theta}$
            \ForEach {$v \in vals$}
                \State $agg \gets agg \oplus v$
            \EndFor
            \State \Return $agg$
        \EndFunction

@@

\begin{algorithm}
    \caption{Subtract-On-Evict insert, evict and query methods}
    \label{pseudo:soe}
    \begin{algorithmic}[1]
        \Function{query}{}

@@

As it turns out, besides \textit{invertibility}, other properties allow us to group aggregations in different families. Tangwongsan et al. construct Table \ref{tbl:aggregations-properties} to summarize the aggregation families and satisfying properties \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.

To understand these properties, we must be familiar with the \textit{combine} and \textit{query} functions. The already mentioned \textit{combine} function merges the incoming event with the incrementally computed aggregation state. The \textit{query} function is responsible for returning a result for a given query according to the current aggregation state. For instance, consider once again the use case of computing a sum aggregation over a window of data. The \textit{combine} function would be the arithmetic operator \textit{+} and would merge the current aggregation state --- \textit{i.e.} the current sum value --- with the new incoming event. The \textit{query} function would return the current aggregation state --- \textit{i.e.} return the total sum computed.

@@

    \item  a \textbf{\textit{unary query}} if the query does not need any parameters --- \textit{e.g.} a membership test with a Bloom filter would require the element to test membership for as a parameter so it is not unary; returning the total sum is unary since it does not require any parameters
\end{enumerate}

@@

According to these properties, Tangwongsan et al. define five aggregation families \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}. These are the sum, collect, median, max and sketch-like families. The sum-like family of aggregations comprises aggregations that have an invertible, associative, commutative and size-preserving \textit{combine} as well as a unary \textit{query} function. Examples of such aggregations are the sum, count, mean/average and standard deviation operations. The collect-like family of aggregations has an invertible and associative \textit{combine} function but their \textit{query} function is not always unary. For example, keeping a string concatenation aggregation is invertible and associative but not commutative nor size-preserving. While in the string concatenation case you do not need to pass any arguments to the \textit{query} function, in the case of another collect-like aggregation, for instance, keeping track of the \textit{i-th} youngest element, you need to specify in your query the parameter \textit{i} (\textit{e.g.} the 5th youngest element). Hence, their \textit{query} function is not always unary. The median-like family of aggregations contains for example the percentile operation. For example, the percentile aggregation is invertible, associative and commutative but not size-preserving and requires an argument to the \textit{query} function: the percentile to compute (\textit{e.g.} percentile 95 or 99). The max-like family comprises aggregations like the maximum or minimum, operations that are not invertible but associative and size-preserving, as well as unary from a \textit{query} function perspective. The final family, the sketch-like family, comprises aggregations that are usually solved by approximate aggregators, also called probabilistic data structures or \textit{sketches}. These aggregations can be membership queries, item frequency estimation or cardinality estimation.

The aggregation properties analyzed in this Section will be used to evaluate SWAG algorithms. Different algorithms might only apply to certain families of aggregations. This way, the multiple SWAG algorithms presented throughout this Thesis will be analyzed not only on their time and space complexity but also on the restrictions they impose regarding the aggregation' properties required. For example, under this framework, to implement Subtract-On-Evict (SOE), both the \textit{combine} $\oplus$ and its inverse $\ominus$ functions must be implemented so that SOE applies $\ominus$ to expire the evicted element from the aggregation state. Since SOE only restriction is for the aggregation to be invertible it works for the \textit{sum-like}, \textit{collect-like} and \textit{median-like} families in Table \ref{tbl:aggregations-properties}.

@@

Definition \ref{def:ema} introduces the recurrence formula for time-based EMAs whereas Definition \ref{def:tuple-ema} works for tuple-based ones. In this Thesis, we chose to work with tuple-based EMAs due to its simplicity, since it does not require keeping track of timestamp variations, but our approach can be adjusted to use time-based ones.

>>>>

Batch processing is the processing of data grouped by batches. A batch is a collection of data points that have been grouped by certain criteria. Examples of such criteria could be \textit{"all the transactions from the past two months"}, \textit{"the last million transactions"} or even \textit{"all the transactions made by an user until now"}. Batch datasets are known as static data \cite{Martin-Batch-Defin}. An example application of such a technique would be processing all the keywords searched for in the google search engine, and then grouping the results by keyword to count how many times a keyword was used in searches (for trend analysis purposes). Between batch and stream processing, batch processing is the older paradigm. Historically, Apache Hadoop \cite{borthakur2007hadoop, Hadoop, ApacheHadoop} was one of the first widely known frameworks for large-scale batch processing, created in 2005. Hadoop uses a programming model called MapReduce \cite{MapReduce}. This paradigm applies a map function that processes key/value pairs and generates sets of intermediate ones, ultimately merging all intermediate values
using a reducer function. To do this, the distributed computation framework reads and writes from disk on intermediate steps, which makes it inefficient for applications that reuse working sets of data across multiple parallel operations (\textit{e.g.}, iterative Machine Learning algorithms). Nearly a decade later, Apache Spark \cite{ApacheSpark, Spark} was released as the industry accepted successor.
Spark retains the scalability and fault tolerance of MapReduce but introduces Resilient Distributed Datasets (RDDs) \cite{SparkRDDs}. An RDD is a read-only collection of objects partitioned across a set of machines that can be rebuilt if a partition is lost.
Additionally, Spark processes data in random access memory (RAM), while Hadoop MapReduce persists data to the disk after a map or reduce action. Thus, Spark achieves lower processing times.

@@

% boundless aggregations no need for windows
Computations over an unbounded dataset are possible. For example, computing the \textit{average} of an event's field for all incoming events is computationally trivial: we need only to store the total number of events and the current total sum of all events arriving in the system. However, is this boundless average over the stream of events useful? While in some scenarios it may be, more often than not we would like to introduce some boundaries into our computations. For instance, consider the monitoring of machines spread out across multiple racks in a large data center and its temperatures. Assume that temperatures measured across all racks are sent to a monitoring system in the form of a data stream. The average temperature of a machine over an infinite period (\textit{i.e.}, since the machine was deployed) is of little value for detecting if the equipment is currently overheating, and to trigger the appropriate action to prevent a failure. In this scenario, an average of the temperatures per rack and for the last 5 minutes allows for a more actionable context, in this case identifying a likely-to-fail rack. To obtain such insight requires computing the average over the last 5 minutes' worth of data, effectively defining a begin and end --- \textit{i.e.}, all the data between the current timestamp in minutes \textit{t} and \textit{t-5}. By creating such a conceptual range, we realize that in practice we need a bounded dataset. Partitioning an infinite data stream to obtain such a finite dataset is made through the use of windows.

Botan \emph{et al.} \cite{Botan-SECRET} define a window as:
\begin{definition}
A window W over a stream $\mathbb{S}$ is a finite subset of $\mathbb{S}$
\end{definition}
Windows can be categorized by two dimensions: the window size $W$ and the sliding step size $S$. Windows can be defined as tuple or time-based. Tuple-based windows are windows whose size is defined by \emph{number} of items contained in the window, whereas time-based windows are windows whose size is defined by the total \emph{time} a window can store. For instance, a tuple-based window of size $10^3$ tuples will store exactly $10^3$ items. In contrast, a time-based window storing the last 5 minutes of events can contain a varying number of tuples but will always keep the events arrived in the last 5 minutes. A window slides over incoming data and groups it, enabling us to compute aggregations over a defined and bounded set of events. The step size \textit{S} of a windowing technique is defined as the distance between consecutive windows \cite{Botan-SECRET}. This means that two consecutive windows W\textsubscript{1} and W\textsubscript{2} are exactly \textit{S} units away. As mentioned, these units can be units of time or a number of tuples.

Windows can also be characterized by how they move across data, \textit{i.e.}, how the step $S$ is defined. The three most common ways are: true sliding, stepping or tumbling windows. A true sliding window is a window with a unitary step, \textit{i.e.}, where $S = 1$. At each step, the window advances one unit, which may be one event for tuple-based steps, or one time-unit whether that is in milliseconds, seconds, days, weeks or another magnitude of time, measured based on the system clock or event time. Such a true sliding movement can be observed in Figure \ref{fig:sliding-window} with a tuple-based window of size \textit{W} of three tuples and unitary step.

@@

For the scope of this thesis, we will work under a tuple-based true sliding window framework. However, despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and other step values as well.

@@

While some aggregations can be computed on an infinite dataset, often we want to compute these over a finite one. As we have seen in Section \ref{sec:windows}, we can obtain a bounded dataset from a data stream by applying windowing techniques. When doing so, we obtain a finite dataset denoted as a sliding window and compute aggregations over it. Tangwongsan \emph{et al.} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs) \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}. As an example of a SWAG, consider the data center temperature monitoring problem from Section \ref{sec:windows}. In this example, we wanted to compute the average temperature per rack in the data center in the last 5 minutes. If we translate this SWAG into streaming SQL following a similar syntax to the one provided by Apache Flink \cite{ApacheFlink} we would get:

@@

In the query above, the
\texttt{TUMBLE(timestamp, Time.minutes(5), rack\_id)} statement groups incoming data using time-based tumbling windows of 5 minutes duration, using the \texttt{timestamp} field to evict events not in this time interval. The resulting list of rows has \texttt{rack\_id} as its key and as corresponding value a list of temperatures, from the past 5 minutes. Finally, for each rack and its list of temperatures, we compute an \texttt{AVG} aggregation, resulting in the average temperature per rack in the past 5 minutes. Note that this is very similar to any aggregation made in a standard Database Management System (DBMS). The main difference is that in standard SQL the \texttt{GROUP BY} clause includes only the columns on which the data will be grouped while in streaming SQL the \texttt{GROUP BY} also specifies the type of window used (\textit{e.g.}, in this case, a tumbling window).

@@

\begin{algorithm}
    \caption[Recalculate-From-Scratch]{Recalculate-From-Scratch insert, evict and query methods}
    \label{pseudo:rfs}
    \begin{algorithmic}[1]
        \Function{query}{vals}
            \State $agg\gets\overline{\theta}$
            \ForEach {$v \in vals$}
                \State $agg \gets agg \oplus v$
            \EndFor
            \State \Return $agg$
        \EndFunction

@@

\begin{algorithm}
    \caption[Subtract-On-Evict]{Subtract-On-Evict insert, evict and query methods}
    \label{pseudo:soe}
    \begin{algorithmic}[1]
        \Function{query}{}

@@

As it turns out, besides \textit{invertibility}, other properties allow us to group aggregations in different families. Tangwongsan \emph{et al.} construct Table \ref{tbl:aggregations-properties} to summarize the aggregation families and satisfying properties \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.

To understand these properties, we must be familiar with the \textit{combine} and \textit{query} functions. The already mentioned \textit{combine} function merges the incoming event with the incrementally computed aggregation state. The \textit{query} function is responsible for returning a result for a given query according to the current aggregation state. For instance, consider once again the use case of computing a sum aggregation over a window of data. The \textit{combine} function would be the arithmetic operator \textit{+} and would merge the current aggregation state --- \textit{i.e.}, the current sum value --- with the new incoming event. The \textit{query} function would return the current aggregation state --- \textit{i.e.}, return the total sum computed.

@@

    \item  a \textbf{\textit{unary query}} if the query does not need any parameters --- \textit{e.g.}, a membership test with a Bloom filter would require the element to test membership for as a parameter so it is not unary; returning the total sum is unary since it does not require any parameters
\end{enumerate}

@@

According to these properties, Tangwongsan \emph{et al.} define five aggregation families \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}. These are the sum, collect, median, max and sketch-like families. The sum-like family of aggregations comprises aggregations that have an invertible, associative, commutative and size-preserving \textit{combine} as well as a unary \textit{query} function. Examples of such aggregations are the sum, count, mean/average and standard deviation operations. The collect-like family of aggregations has an invertible and associative \textit{combine} function but their \textit{query} function is not always unary. For example, keeping a string concatenation aggregation is invertible and associative but not commutative nor size-preserving. While in the string concatenation case you do not need to pass any arguments to the \textit{query} function, in the case of another collect-like aggregation, for instance, keeping track of the \textit{i-th} youngest element, you need to specify in your query the parameter \textit{i} (\textit{e.g.}, the 5th youngest element). Hence, their \textit{query} function is not always unary. The median-like family of aggregations contains for example the percentile operation. For example, the percentile aggregation is invertible, associative and commutative but not size-preserving and requires an argument to the \textit{query} function: the percentile to compute (\textit{e.g.}, percentile 95 or 99). The max-like family comprises aggregations like the maximum or minimum, operations that are not invertible but associative and size-preserving, as well as unary from a \textit{query} function perspective. The final family, the sketch-like family, comprises aggregations that are usually solved by approximate aggregators, also called probabilistic data structures or \textit{sketches}. These aggregations can be membership queries, item frequency estimation or cardinality estimation.

The aggregation properties analyzed in this section will be used to evaluate SWAG algorithms. Different algorithms might only apply to certain families of aggregations. This way, the multiple SWAG algorithms presented throughout this thesis will be analyzed not only on their time and space complexity but also on the restrictions they impose regarding the aggregation' properties required. For example, under this framework, to implement Subtract-On-Evict (SOE), both the \textit{combine} $\oplus$ and its inverse $\ominus$ functions must be implemented so that SOE applies $\ominus$ to expire the evicted element from the aggregation state. Since SOE only restriction is for the aggregation to be invertible it works for the \textit{sum-like}, \textit{collect-like} and \textit{median-like} families in Table \ref{tbl:aggregations-properties}.

@@

Definition \ref{def:ema} introduces the recurrence formula for time-based EMAs whereas Definition \ref{def:tuple-ema} works for tuple-based ones. In this thesis, we chose to work with tuple-based EMAs due to its simplicity, since it does not require keeping track of timestamp variations, but our approach can be adjusted to use time-based ones.

### 2020-07-05 chapters/introduction.tex
Multiple grammar corrections

\section{Context} \label{sec:context}
In the past decade, applications have become increasingly data-driven, placing data at the center of application design. For different use cases, the data is processed in different ways for different purposes. For instance, e-commerce platforms such as Alibaba\footnote{https://alibaba.com} need to process a large number of daily transactions while ensuring that sales run smoothly and that products are delivered to customers' homes. Streaming services, like YouTube\footnote{https://youtube.com} and Netflix\footnote{https://netflix.com}, try to guarantee that media content reaches up to millions of users simultaneously. Social networks are responsible for generating large volumes of data of which,  Twitter\footnote{https://twitter.com} for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day\footnote{https://blog.twitter.com/engineering/en\_us/a/2013/new-tweets-per-second-record-and-how.html}. \textit{Tweets} may contain text, media content or both. Cybersecurity applications are yet another example of time-sensitive, data-driven applications. The entities that work in these use-cases need to monitor user accesses and their actions in the network, where timely detection of intruders is critical to prevent them from tampering with the underlying system.

All of these applications generate large volumes of data over time which poses data storage and processing challenges. This data is characterized by its large volume, high velocity and high variety \cite{Mavragani-GoogleTrends-SLR}, creating large scale data management problems. Additionally, the large amounts of information produced by such systems and use-cases lead to the creation of multiple real-time unbounded data sets, also known as data streams. The information that flows through such a stream can be analyzed on-the-fly or stored for later processing. The former is best known as stream processing, whereas the latter is known as batch processing.

@@

As mentioned, the processing of data can be divided into two main categories: batch and stream processing. However, due to the increasing volume of data and the need for timely processing to allow companies to react to changing conditions in real-time, the demand for stream processing systems is increasing. Unlike batch processing of static data sets, where assumptions on the underlying data distributions can be made, streaming data is known for being non-stationary \cite{Gama-Knowledge-Discovery} where the value of the produced information lies in its recency \cite{Kolajo-Big-data-stream-SLR}. \textit{Recency} is measured under different scales for different use cases. For instance, consider the monitoring of geological data to forecast possible natural disasters such as earthquakes versus the monitoring of a computer network for intruder detection. In the former case, after the forecast of a future earthquake, governmental authorities need a couple of days to launch an evacuation plan and keep the population safe. However, in the intruder detection scenario, the decision of removing access to a user must be done as soon as possible, preferably in a few seconds. Hence, while in the first scenario information retrieved within a day would still be recent, in the second scenario, information is considered recent and relevant if delivered within seconds (or even milliseconds).

In either processing scenario, we usually want to apply operations on data, for example, to compute the maximum value, an average or a count of distinct elements. Operations that produce a single result when applied to data sets are commonly known as aggregation operations. In stream processing, we apply these aggregations to data streams to obtain valuable information. However, since a data stream is, effectively speaking, an unbounded data set, if our aggregation requires a finite domain, there is the need to apply windowing techniques. A thorough analysis of stream windowing techniques is done in Chapter \ref{chap:background} of this document but for now it suffices to say that windowing is a mechanism to select a fine portion of an unbounded data stream, typically by defining a time or tuple-based window size --- \textit{e.g.}, events from the last three days or the last million events, respectively.

Consider the use-case where we want to detect credit card fraud on the data stream of all the transactions made on Amazon\footnote{https://amazon.com} by every US citizen on a single day. In this case, Amazon's transaction stream will need to be monitored by another system, a fraud detection one. Such a system would have to process all incoming transactions and decide, in a fraction of a second, if a transaction is fraudulent or not. It is clear that manually monitoring such an intensive, latency-sensitive data stream is hard and error-prone, thus justifying the need for autonomous real-time data stream monitoring systems.

Many market solutions monitor streams of credit card transactions with the intent of fraud detection. In the context of this Thesis, we work side by side with Feedzai\footnote{https://feedzai.com}. In brief, the company's solution for fraud detection consists of a workflow of different components. Incoming events are fed into this workflow and processed accordingly. Workflow components can be sets of user-defined rules and/or Machine Learning (ML) models that output scores reflecting their belief on whether a transaction is fraudulent or not. ML models are configured and trained,  before deployment, often with assumptions on the statistical distributions of future incoming data. Thus their model performance depends on how well those assumptions hold for future data. However, fraud patterns are not static over time, but seasonal --- \textit{e.g.}, Black Fridays, holiday shopping and product launches --- and evolutionary --- \textit{e.g.}, new fraud attack methods by fraudsters. Due to these shifts in data patterns and distributions, the ML models' performance can deteriorate over periods of time. In the Machine Learning field, this performance decay caused by data pattern shifts is also known as concept drift \cite{drift, gama-drift}.

@@

We want to build a lightweight enough solution that can be integrated into existing workflows. Ensuring constant time and space complexity means that the system will remain usable for the ever-growing volumes of data. Since such a monitoring system is not mission-critical --- \textit{i.e.} the provided services do not rely on it --- the operational cost associated with it must be kept to a minimum. Having a real-time system monitoring other real-time systems consuming as many resources as the latter would be far too costly.

@@

    \item explainable alerts
\end{itemize}
The hypothesis will be tested and evaluated on the previous criteria. Experiments will be made using multiple synthetic data sets and a real dataset from the financial fraud space.

>>>>

\section{Context} \label{sec:context}
In the past decade, applications have become increasingly data-driven, placing data at the center of application design. For different use cases, the data is processed in different ways for different purposes. For instance, e-commerce platforms such as Alibaba\footnote{https://alibaba.com} need to process a large number of daily transactions while ensuring that sales run smoothly and that products are delivered to customers' homes. Streaming services, like YouTube\footnote{https://youtube.com} and Netflix\footnote{https://netflix.com}, try to guarantee that media content reaches up to millions of users simultaneously. Social networks are responsible for generating large volumes of data of which,  Twitter\footnote{https://twitter.com} for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day\footnote{https://blog.twitter.com/engineering/en\_us/a/2013/new-tweets-per-second-record-and-how.html}. \textit{Tweets} may contain text, media content or both. Cybersecurity applications are yet another example of time-sensitive, data-driven applications. The entities that work in these use cases need to monitor user accesses and their actions in the network, where timely detection of intruders is critical to prevent them from tampering with the underlying system.

All of these applications generate large volumes of data over time which poses data storage and processing challenges. This data is characterized by its large volume, high velocity and high variety \cite{Mavragani-GoogleTrends-SLR}, creating large scale data management problems. Additionally, the large amounts of information produced by such systems and use cases lead to the creation of multiple real-time unbounded datasets, also known as data streams. The information that flows through such a stream can be analyzed on-the-fly or stored for later processing. The former is best known as stream processing, whereas the latter is known as batch processing.

@@

As mentioned, the processing of data can be divided into two main categories: batch and stream processing. However, due to the increasing volume of data and the need for timely processing to allow companies to react to changing conditions in real-time, the demand for stream processing systems is increasing. Unlike batch processing of static datasets, where assumptions on the underlying data distributions can be made, streaming data is known for being non-stationary \cite{Gama-Knowledge-Discovery} where the value of the produced information lies in its recency \cite{Kolajo-Big-data-stream-SLR}. \textit{Recency} is measured under different scales for different use cases. For instance, consider the monitoring of geological data to forecast possible natural disasters such as earthquakes versus the monitoring of a computer network for intruder detection. In the former case, after the forecast of a future earthquake, governmental authorities need a couple of days to launch an evacuation plan and keep the population safe. However, in the intruder detection scenario, the decision of removing access to a user must be done as soon as possible, preferably in a few seconds. Hence, while in the first scenario information retrieved within a day would still be recent, in the second scenario, information is considered recent and relevant if delivered within seconds (or even milliseconds).

In either processing scenario, we usually want to apply operations on data, for example, to compute the maximum value, an average or a count of distinct elements. Operations that produce a single result when applied to datasets are commonly known as aggregation operations. In stream processing, we apply these aggregations to data streams to obtain valuable information. However, since a data stream is, effectively speaking, an unbounded dataset, if our aggregation requires a finite domain, there is the need to apply windowing techniques. A thorough analysis of stream windowing techniques is done in Chapter \ref{chap:background} of this document but for now it suffices to say that windowing is a mechanism to select a fine portion of an unbounded data stream, typically by defining a time or tuple-based window size --- \textit{e.g.}, events from the last three days or the last million events, respectively.

Consider the use case where we want to detect credit card fraud on the data stream of all the transactions made on Amazon\footnote{https://amazon.com} by every US citizen on a single day. In this case, Amazon's transaction stream will need to be monitored by another system, a fraud detection one. Such a system would have to process all incoming transactions and decide, in a fraction of a second, if a transaction is fraudulent or not. It is clear that manually monitoring such an intensive, latency-sensitive data stream is hard and error-prone, thus justifying the need for autonomous real-time data stream monitoring systems.

Many market solutions monitor streams of credit card transactions with the intent of fraud detection. In the context of this thesis, we work side by side with Feedzai\footnote{https://feedzai.com}. In brief, the company's solution for fraud detection consists of a workflow of different components. Incoming events are fed into this workflow and processed accordingly. Workflow components can be sets of user-defined rules and/or Machine Learning (ML) models that output scores reflecting their belief on whether a transaction is fraudulent or not. ML models are configured and trained,  before deployment, often with assumptions on the statistical distributions of future incoming data. Thus their model performance depends on how well those assumptions hold for future data. However, fraud patterns are not static over time, but seasonal --- \textit{e.g.}, Black Fridays, holiday shopping and product launches --- and evolutionary --- \textit{e.g.}, new fraud attack methods by fraudsters. Due to these shifts in data patterns and distributions, the ML models' performance can deteriorate over periods of time. In the Machine Learning field, this performance decay caused by data pattern shifts is also known as concept drift \cite{drift, gama-drift}.

@@

We want to build a lightweight enough solution that can be integrated into existing workflows. Ensuring constant time and space complexity means that the system will remain usable for the ever-growing volumes of data. Since such a monitoring system is not mission-critical --- \textit{i.e.}, the provided services do not rely on it --- the operational cost associated with it must be kept to a minimum. Having a real-time system monitoring other real-time systems consuming as many resources as the latter would be far too costly.

@@

    \item explainable alerts
\end{itemize}
The hypothesis will be tested and evaluated on the previous criteria. Experiments will be made using multiple synthetic datasets and a real dataset from the financial fraud space.

### 2020-07-05 chapters/method.tex
Multiple grammar corrections

\subsection*{Challenges}
The analyzed outlier detection algorithms in Section \ref{sec:outliers} were deemed unfit for our use-case due to one or more of the following: \textit{(a)} high memory consumption, \textit{(b)} high time complexity, \textit{(c)} "memory loss" and \textit{(d)} obfuscated insights.

For this Thesis, we need constant time complexity to process events one by one in real-time. Because our system is not mission-critical, we also want the solution to be as lightweight as possible, which means techniques that grow linearly with window size are not a good fit. Some of the methods discussed also suffer from what we call \textit{"memory loss"}: they lose track of the reference window or normality period. In other words, the algorithms are online learners, meaning they adapt to the underlying statistics of the data stream. While in some scenarios this is a much-needed feature, in our use-case it is not. How clear the alerts and insights retrieved from the methods are, is also an important property. Some outlier detection methods use Machine Learning (ML) techniques to produce alerts, which makes it harder to explain to a system administrator why an alert was raised and what changed.

@@

In Section \ref{sec:stream-superset}, we defined a data stream as a continuous collection of events. Each event is timestamped and contains multiple fields. In the context of this Thesis, each of the event's fields are named features. Our goal is to detect data pattern shifts for each feature in a streaming fashion. Such shifts will be measured between a static in time or reference period versus the observed reality during the streaming phase. Figure \ref{fig:timelines} shows the separated reference and streaming timelines.

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.7]{figures/compute-sample-distances.png}
      \caption[Compute as many distance values as samples]{Compute as many distance values as samples, one for each sample-reference histogram pair}
      \label{fig:compute-sample-distances}
    \end{center}
\end{figure}

>>>>

\subsection*{Challenges}
The analyzed outlier detection algorithms in Section \ref{sec:outliers} were deemed unfit for our use case due to one or more of the following: \textit{(a)} high memory consumption, \textit{(b)} high time complexity, \textit{(c)} "memory loss" and \textit{(d)} obfuscated insights.

For this thesis, we need constant time complexity to process events one by one in real-time. Because our system is not mission-critical, we also want the solution to be as lightweight as possible, which means techniques that grow linearly with window size are not a good fit. Some of the methods discussed also suffer from what we call \textit{"memory loss"}: they lose track of the reference window or normality period. In other words, the algorithms are online learners, meaning they adapt to the underlying statistics of the data stream. While in some scenarios this is a much-needed feature, in our use case it is not. How clear the alerts and insights retrieved from the methods are, is also an important property. Some outlier detection methods use Machine Learning (ML) techniques to produce alerts, which makes it harder to explain to a system administrator why an alert was raised and what changed.

@@

In Section \ref{sec:stream-superset}, we defined a data stream as a continuous collection of events. Each event is timestamped and contains multiple fields. In the context of this thesis, each of the event's fields are named features. Our goal is to detect data pattern shifts for each feature in a streaming fashion. Such shifts will be measured between a static in time or reference period versus the observed reality during the streaming phase. Figure \ref{fig:timelines} shows the separated reference and streaming timelines.

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.7]{figures/compute-sample-distances.png}
      \caption[Compute sample's distance values]{Compute as many distance values as samples, one for each sample-reference histogram pair}
      \label{fig:compute-sample-distances}
    \end{center}
\end{figure}

### 2020-07-05 chapters/problem-statement.tex
Multiple grammar corrections

\section{State of the Art Issues} \label{sec:sotaissues}
Aggregations, specifically sliding window ones, were introduced in Chapter \ref{chap:background} along with several windowing techniques required to monitor streaming data patterns in real-time. In Chapter \ref{chap:sota} we presented state of the art sliding window aggregation algorithms as well as outlier detection methods. However, the reviewed methods were deemed unfit for our use-case because none of them simultaneously showed all of the desired characteristics, namely:

@@

\section{Research Questions} \label{sec:rqs}
The goal of this Thesis is to develop a lightweight stream monitoring system that detects changes in features distribution between a reference period and a sliding window in real-time. To that end, we devise a series of research questions to be answered by the end of this Thesis:

>>>>

\section{State of the Art Issues} \label{sec:sotaissues}
Aggregations, specifically sliding window ones, were introduced in Chapter \ref{chap:background} along with several windowing techniques required to monitor streaming data patterns in real-time. In Chapter \ref{chap:sota} we presented state of the art sliding window aggregation algorithms as well as outlier detection methods. However, the reviewed methods were deemed unfit for our use case because none of them simultaneously showed all of the desired characteristics, namely:

@@

\section{Research Questions} \label{sec:rqs}
The goal of this thesis is to develop a lightweight stream monitoring system that detects changes in features distribution between a reference period and a sliding window in real-time. To that end, we devise a series of research questions to be answered by the end of this thesis:
